---
title: "Research project"
author: "Oleksii Lasiichuk, Denys Marchenko, Marko Pasternak"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

# Probability and Statistics

# Research project: Statistical analysys of chess games

------------------------------------------------------------------------

## Short desription

This research project utilizes a dataset of chess games played on the
Lichess platform to investigate statistical patterns in online play. Our
analysis focuses on two primary objectives: first, to determine if the
classification of an opening (Open, Closed, Semi-Open, Semi-Closed, or
Flank) significantly influences the duration of a game; and second, to
evaluate the predictive power of rating differentials in determining
match outcomes.

------------------------------------------------------------------------

## Data Preprocessing

The [dataset](https://www.kaggle.com/datasets/datasnaek/chess) utilized
in this study comprises approximately 20,000 chess games played online
on the Lichess platform. The raw dataset contains the following
variables:

-   **Game ID:** Unique identifier for the match.
-   **Rated (T/F):** Boolean indicating if the match affects player
    ratings.
-   **Start Time / End Time:** Timestamp of the game.
-   **Number of Turns:** Total moves played.
-   **Game Status:** Result condition (e.g., resign, mate, draw).
-   **Winner:** White, Black, or Draw.
-   **Time Increment:** Time control settings.
-   **White/Black Player ID:** Usernames of the players.
-   **White/Black Player Rating:** ELO rating of the players.
-   **Moves:** Full move history in standard chess notation.
-   **Opening Eco:** Standardized Encyclopedia of Chess Openings code.
-   **Opening Name:** The traditional name of the opening.
-   **Opening Ply:** Number of moves in the opening phase.

### Feature Selection and Engineering

To facilitate our specific research aims, we performed data cleaning and
feature engineering.

First, we removed metadata columns that are unimportant to our
statistical analysis of game mechanics. Specifically, **Game ID**,
**Start Time**, **End Time**, **White Player ID**, and **Black Player
ID** were excluded.

Second, we removed all unrated games. Since these matches are typically
used for training or casual play, they were treated as
non-representative outliers to prevent skewing the results.

Third, we engineered several new variables to categorize the data for
hypothesis testing:

1.  **Opening Type:** A categorical variable classifying the opening
    into one of five standard chess systems: *Open*, *Closed*,
    *Semi-Open*, *Semi-Closed*, or *Flank*.
2.  **Average Rating:** The mean rating of the two players
    `(White Rating + Black Rating) / 2`, used to gauge the overall skill
    level of the match.
3.  **Rating Gap:** The absolute difference in rating between the two
    players (`abs(White Rating - Black Rating)`).
4.  **Outcome:** A categorical variable describing the result relative
    to rating expectations:
    -   *Expected Win:* The higher-rated player won.
    -   *Upset:* The lower-rated player won.
    -   *Draw:* The game ended in a draw.
    -   *Even:* Both players have the same rating.
5.  **Gap Category:** A binned version of the Rating Gap variable (e.g.,
    0–50, 50–100, 100–150) to allow for grouped analysis of rating
    disparities.

```{r}
library(dplyr)
library(stringr)

# Load data
chess_data <- read.csv("games.csv")

clean_data <- chess_data %>%
  # 1. Filter for RATED games only
  filter(rated == TRUE) %>%
  
  # 2. Feature Selection (Remove columns)
  select(-id, -created_at, -last_move_at, -white_id, -black_id, -rated) %>%
  
  # 2. Feature Engineering
  mutate(
    # Average Rating
    average_rating = (white_rating + black_rating) / 2,
    
    # Rating Gap (Absolute difference)
    rating_gap = abs(white_rating - black_rating),
    
    # Opening Type
    opening_first_two = word(moves, 1, 2), # Extract first two moves
    opening_type = case_when(
      opening_first_two == "e4 e5" ~ "Open Game",
      str_detect(opening_first_two, "^e4") & opening_first_two != "e4 e5" ~ "Semi-Open Game",
      opening_first_two == "d4 d5" ~ "Closed Game",
      str_detect(opening_first_two, "^d4") & opening_first_two != "d4 d5" ~ "Semi-Closed Game",
      TRUE ~ "Flank Opening"
    ),
    
    # Outcome (Expected vs Upset)
    outcome_type = case_when(
      winner == "draw" ~ "Draw",
      (winner == "white" & white_rating > black_rating) ~ "Expected Win",
      (winner == "black" & black_rating > white_rating) ~ "Expected Win",
      TRUE ~ "Upset"
    ),
    
    # Gap Category (Binning the gap)
    gap_category = cut(rating_gap, 
                       breaks = seq(0, max(rating_gap)+50, by = 50), 
                       include.lowest = TRUE)
  )


head(clean_data)

```

------------------------------------------------------------------------

## The Impact of Opening Type on Game Duration

### Classification of Chess Openings

The opening phase is critical in determining the trajectory and
character of a chess game. We categorize games into five distinct types
based on the initial move order:

-   **Open Games (**$1. e4\ e5$): Characterized by immediate interaction
    in the center. These positions often lead to open lines for piece
    activity, favoring tactical and dynamic play.
-   **Closed Games (**$1. d4\ d5$): Characterized by a stabilized
    center. These games often evolve into slower, strategic battles
    where maneuvering on the queenside is more common than direct
    attacks on the king.
-   **Semi-Open Games (**$1. e4$, black responds $\neq e5$): Black
    allows White to occupy the center to attack it later. This creates
    immediate structural imbalances, often resulting in sharp,
    counter-attacking games (e.g., the Sicilian Defense).
-   **Semi-Closed Games (**$1. d4$, black responds $\neq d5$): Black
    responds asymmetrically to $d4$. These systems lead to complex
    strategic imbalances where predicting the flow is difficult.
-   **Flank Openings (White moves** $\neq e4, d4$): White attacks the
    center from the sides (e.g., $1. c4$ or $1. Nf3$). These are highly
    flexible and hard to generalize.

*Note:* While these classifications describe statistical tendencies,
they are not deterministic. Open games can transpose into quiet
positions, and Closed games can become tactically volatile.

### Hypothesis Formulation

We aim to test whether the chosen opening category has a statistically
significant effect on the duration of the game (measured in number of
moves).

Let $D$ be the random variable representing the duration of a game. We
define the median duration for each opening type as follows:

-   $\theta_{open}$: Median moves for Open games

-   $\theta_{closed}$: Median moves for Closed games

-   $\theta_{semi\_open}$: Median moves for Semi-Open games

-   $\theta_{semi\_closed}$: Median moves for Semi-Closed games

-   $\theta_{flank}$: Median moves for Flank games

**Null Hypothesis (**$H_0$): The type of opening has no effect on game
duration. The median game lengths are equal across all categories.
$$H_0: \theta_{open} = \theta_{closed} = \theta_{semi\_open} = \theta_{semi\_closed} = \theta_{flank}$$

**Alternative Hypothesis (**$H_1$): The opening type influences the game
duration. At least one group's median differs from the others.
$$H_1: \exists (i, j) \text{ such that } \theta_i \neq \theta_j$$

**Theoretical Prediction:** Based on the tactical nature of **Open
Games**, we predict $\theta_{open}$ will be the lowest, as tactical
errors often lead to quick decisive results (checkmate or resignation).
Conversely, we predict $\theta_{closed}$ will be the highest, as these
games rely on long-term positional maneuvering where decisive advantages
take longer to accumulate.

### Methodology

To test these hypotheses, we will utilize the **Kruskal-Wallis Rank Sum
Test**.

This non-parametric test is expedient for our analysis because:

1.  **Non-Normality:** We anticipate that the distribution of chess game
    lengths will be right-skewed and non-normal. This would violate the
    normality assumption required for a standard **ANOVA** (Analysis of
    Variance), which is the traditional parametric test used to compare
    the means of three or more groups. We will verify this distribution
    visually in the Descriptive Analysis section.

2.  **Multiple Groups:** We are comparing five distinct independent
    groups. As the **Mann-Whitney U test** is limited to comparing the
    distributions of only *two* groups, the Kruskal-Wallis test serves
    as the necessary extension to handle our multi-group comparison
    simultaneously.

If the Kruskal-Wallis test indicates significance ($p < 0.05$), we will
reject $H_0$ and conclude that opening choice does impact game
longevity.

### Descriptive analysis of data

```{r}
library(ggplot2)
library(dplyr)
library(knitr)

# --- 1. Summary Statistics Table (Unchanged) ---
desc_stats <- clean_data %>%
  group_by(opening_type) %>%
  summarise(
    Count = n(),
    Mean_Turns = mean(turns, na.rm = TRUE),
    Median_Turns = median(turns, na.rm = TRUE),
    SD_Turns = sd(turns, na.rm = TRUE),
    Min = min(turns, na.rm = TRUE),
    Max = max(turns, na.rm = TRUE)
  ) %>%
  arrange(desc(Median_Turns))

kable(desc_stats, caption = "Summary Statistics of Game Duration by Opening Type")

# --- 2. Individual Histograms Loop ---
# We get a list of all unique opening types
types <- unique(clean_data$opening_type)

# We loop through each type to create a separate graph
for (type in types) {
  
  # Filter data for just this one opening type
  subset_data <- clean_data %>% filter(opening_type == type)
  
  # Calculate the median for this specific group to draw a vertical line
  this_median <- median(subset_data$turns, na.rm = TRUE)
  
  # Create the plot
  p <- ggplot(subset_data, aes(x = turns)) +
    geom_histogram(aes(y = after_stat(density)), binwidth = 5, fill = "#69b3a2", color = "white", alpha = 0.8) +
    geom_density(alpha = 0.2, fill = "#69b3a2", linewidth = 1) +
    
    # Add a vertical red line for the Median
    geom_vline(aes(xintercept = this_median), color = "red", linetype = "dashed", linewidth = 1) +
    
    # Add a text label for the median value
    annotate("text", x = this_median + 10, y = 0.01, label = paste("Median:", this_median), color = "red") +
    
    labs(
      title = paste("Distribution of Duration:", type),
      subtitle = paste("Based on", nrow(subset_data), "games"),
      x = "Number of Turns",
      y = "Density"
    ) +
    theme_minimal() +
    coord_cartesian(xlim = c(0, 150))
  
  # In R loops, you must explicitly call print() to show the graph
  print(p)
}
```

**1. Distribution Shape:** Visual inspection of the histograms confirms
that game duration is **heavily right-skewed** across all five opening
categories. The "long tail" extending to the right indicates that while
most games finish between 40-60 moves, outliers of 100+ moves are
common. This non-normality validates our decision to use the median as
our central metric and the non-parametric **Kruskal-Wallis test** for
hypothesis testing.

**2. Analysis of Central Tendency:** Our data strongly supports the
hypothesis regarding **Open Games**. With a median of **53 moves**, they
are the shortest category, distinctly separated from the others. This
aligns with the theoretical understanding that $1. e4\ e5$ positions
often lead to sharp, tactical clashes that resolve quickly.

**3. The "Strategic Cluster":** Interestingly, the data contradicts the
assumption that Closed games are uniquely long. Instead, we observe a
"strategic cluster": **Closed (57), Semi-Open (57), and Semi-Closed
(58)** games all share a nearly identical median duration. This suggests
that any deviation from the open $1. e4\ e5$ structure—whether through
$d4$ or asymmetric defenses—tends to extend the game by an average of
4-5 moves.

**4. Flank Openings:** Flank openings occupy a transitional middle
ground with a median of **55 moves**. While visual inspection suggests a
distinct distribution profile, statistical analysis shows their
volatility (Standard Deviation $\approx 33$) is consistent with all
other opening types. This indicates that high variance in game length is
an inherent property of chess itself, rather than a specific feature of
Flank systems.

To characterize the nature of these openings beyond simple duration, we
examine three additional statistics:

-   **Interquartile Range (IQR):** Measures the spread of the middle 50%
    of the data. In this context, IQR serves as a proxy for
    **consistency**; a higher value indicates unpredictable, volatile
    game lengths.
-   **Miniature Rate (**$< 25$ moves): Measures the percentage of games
    ending quickly. This serves as a metric for **sharpness**,
    identifying openings that lead to immediate tactical decisiveness.
-   **Draw Rate:** Measures the percentage of games ending in a draw.
    This correlates with **strategic stability**, as solid positions are
    harder to break down and often extend into long endgames.

```{r}
# Extended Descriptive Statistics Table
extended_stats <- clean_data %>%
  group_by(opening_type) %>%
  summarise(
    Median = median(turns),
    IQR = IQR(turns),
    Miniature_Rate = paste0(round(mean(turns < 25) * 100, 1), "%"),
    Draw_Rate = paste0(round(mean(winner == "draw") * 100, 1), "%")
  ) %>%
  arrange(desc(Median))

kable(extended_stats, caption = "Volatility and Character Metrics by Opening Type")
```

Beyond the median duration, the supplementary metrics reveal distinct
behavioral profiles for each opening category:

**1. The "Sharpness" of Open Games:** The **Miniature Rate** (\< 25
moves) provides the strongest evidence for our hypothesis regarding the
volatility of Open Games. At **13.6%**, the probability of an Open Game
ending in the opening phase is **exactly double** that of a Semi-Closed
Game (6.8%). This statistically confirms that $1. e4\ e5$ positions
carry a significantly higher risk of immediate tactical decisiveness
compared to $1. d4$ systems.

**2. The Unexpected Decisiveness of Flank Openings:** Contrary to the
common belief that Flank openings are passive, the data shows they are
highly decisive. They possess the **lowest draw rate (3.5%)** of all
categories and a **Miniature Rate (13%)** nearly identical to Open
Games. This suggests that while Flank openings (like the English or
Reti) are positionally complex, they create imbalances that force a
result—win or loss—rather than leading to safe, sterile draws.

**3. Stability of Variance (IQR):** The **Interquartile Range (IQR)** is
remarkably consistent across all groups, ranging tightly between **40
and 43 moves**. This is a crucial negative finding: it indicates that
while the *median* length shifts depending on the opening, the
*unpredictability* (spread) of the game length is an inherent property
of chess itself, rather than a feature of any specific opening system.

**4. Strategic Solidity:** The **Semi-Closed** and **Closed** games
exhibit the highest draw rates (5.1% and 4.7%) and the lowest miniature
rates. This inverse relationship—low risk of quick loss combined with
higher draw probability—statistically defines these openings as the most
"solid" choices for risk-averse players.

**Conclusion & Next Steps:** We have visually identified a likely
difference between **Open Games** and the **Rest**. The next step is to
perform the Kruskal-Wallis test to determine if these observed
differences (53 vs. 58) are statistically significant or merely due to
random variance.

```{r}
# Load necessary libraries
library(broom) # For tidying statistical outputs
library(knitr) # For nice tables

# --- 1. The Global Test (Kruskal-Wallis) ---
# This checks if at least one group is different from the others
kw_result <- kruskal.test(turns ~ opening_type, data = clean_data)

print(kw_result)

# --- 2. Interpretation & Post-Hoc Analysis ---
# We only run step 2 if the p-value is significant (< 0.05)
if(kw_result$p.value < 0.05) {
  
  # Pairwise Wilcoxon Test with Bonferroni correction
  # This compares every opening against every other opening
  pairwise_result <- pairwise.wilcox.test(clean_data$turns, 
                                          clean_data$opening_type, 
                                          p.adjust.method = "bonferroni")
  
  # Display the matrix of p-values
  # Values < 0.05 indicate a statistically significant difference between those two specific groups

  print(pairwise_result)
  
} else {
  print("Result: No significant difference found. We cannot reject the Null Hypothesis.")
}
```

### Hypothesis Testing Results

**1. Global Significance (Kruskal-Wallis Test):** The Kruskal-Wallis
rank sum test yielded a p-value of $1.852 \times 10^{-7}$, which is well
below the significance threshold ($\alpha = 0.05$). \* **Conclusion:**
We reject the Null Hypothesis ($H_0$). There is a statistically
significant difference in median game duration between at least two of
the opening categories.

**2. Pairwise Comparisons (Post-Hoc Analysis):** The pairwise Wilcoxon
test (with Bonferroni correction) reveals the specific nature of these
differences:

-   **The "Open Game" Distinction:** Open Games ($1. e4\ e5$) are
    statistically distinct from all "strategic" categories (Closed,
    Semi-Closed, and Semi-Open), with p-values consistently below
    $0.005$. This confirms our descriptive finding that Open games are
    uniquely faster than the norm.

-   **The "Strategic Cluster" Consistency:** Comparing Closed,
    Semi-Closed, and Semi-Open games against each other yields p-values
    of $1.00$ or close to it. This indicates no statistical difference
    between them. Regardless of whether the game starts with $1. d4$ or
    an asymmetric defense, the game duration converges to the same
    "slower" baseline.

-   **The Flank Anomaly:** Interestingly, Flank Openings show no
    statistically significant difference when compared to Open Games
    ($p=1.00$), despite having a slightly higher median (55 vs 53).
    However, they are distinct from Semi-Closed games ($p=0.008$). This
    places Flank openings in a statistical "middle ground"—neither fully
    "sharp" nor fully "slow."

**Final Verdict:** Our data proves that **opening choice is a
significant predictor of game length**. Specifically, choosing an **Open
Game (**$1. e4\ e5$) significantly reduces the expected duration of the
match compared to Closed or Semi-Closed systems.

------------------------------------------------------------------------




## Evaluating Rating Difference as an Outcome Predictor

### Aim and Hypotheses

In this section, we investigate whether the magnitude of the rating difference 
between two players allows us to predict the game outcome. The ELO rating system 
is designed to measure chess skill, so we expect that a larger rating advantage 
should translate to a higher probability of winning for the favorite.

### Definition of Variables

Let $Y$ be a binary random variable representing the game outcome for the 
higher-rated player:

- $Y = 1$: The higher-rated player wins
- $Y = 0$: The higher-rated player does not win (Loss or Draw)

Let $X$ denote the absolute difference in ELO ratings between the two players.

We model the probability $P(Y=1)$ using a **Logistic Regression** model:
$$\ln\left(\frac{P(Y=1)}{1-P(Y=1)}\right) = \beta_0 + \beta_1 X$$

where $\beta_1$ represents the effect of each additional ELO point difference.

### Statistical Hypotheses

**Null Hypothesis** ($H_0$): Rating difference has no effect on win probability.
$$H_0: \beta_1 = 0$$

**Alternative Hypothesis** ($H_1$): Rating difference affects win probability.
$H_1: \beta_1 \neq 0$

We use a significance level of $\alpha = 0.05$ for our hypothesis test.

### Methodology

We chose **Logistic Regression** for several reasons:

1. **Binary Outcome**: Our dependent variable (win/not win) is binary, making 
   logistic regression the appropriate choice.
2. **Probability Modeling**: The logistic function naturally constrains predictions 
   between 0 and 1, which aligns with probability interpretation.
3. **Interpretability**: Coefficients can be transformed into odds ratios, providing 
   intuitive interpretation of effect size.
4. **Robust**: Doesn't assume normal distribution of the predictor variable.

### Data Preparation

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)

# Prepare data for modeling
model_data <- clean_data %>%
  filter(white_rating != black_rating) %>%  # Remove equal ratings
  mutate(
    rating_diff = abs(white_rating - black_rating),
    favorite_won = case_when(
      (white_rating > black_rating) & winner == "white" ~ 1,
      (black_rating > white_rating) & winner == "black" ~ 1,
      TRUE ~ 0
    )
  )
```

After filtering out games where both players have equal ratings, we have **`r nrow(model_data)`** games available for analysis. We excluded **`r nrow(clean_data) - nrow(model_data)`** games where players had identical ratings. The baseline win rate for the higher-rated player is **`r round(mean(model_data$favorite_won) * 100, 1)`%**, which serves as our benchmark—our model needs to provide insights beyond this simple baseline.

### Descriptive Analysis

First, let's examine the distribution of rating differences to understand our data density.
```{r}
# Summary statistics
cat("Mean Rating Diff:", round(mean(model_data$rating_diff), 1), "\n")
cat("Median Rating Diff:", round(median(model_data$rating_diff), 1), "\n")
cat("Max Rating Diff:", max(model_data$rating_diff), "\n")

# Distribution plot
ggplot(model_data, aes(x = rating_diff)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Rating Differences",
    subtitle = "Most games occur between players of similar skill (Right-Skewed)",
    x = "Rating Difference (ELO Points)",
    y = "Count of Games"
  ) +
  theme_minimal()
```

**Interpretation**: 
The distribution is strongly right-skewed. Most matchmaking occurs between players with a rating difference of less than 200 points. This means our model will be most robust in this range, while predictions for extreme rating differences (e.g., >600) rely on fewer data points.

Next, we visualize the relationship between rating difference and win rate:

```{r}
# Calculate win rates by rating difference bins
win_rate_summary <- model_data %>%
  mutate(rating_bin = cut(rating_diff, 
                          breaks = seq(0, max(rating_diff) + 50, by = 50),
                          include.lowest = TRUE)) %>%
  group_by(rating_bin) %>%
  summarise(
    rating_diff_mid = mean(rating_diff),
    win_percentage = mean(favorite_won) * 100,
    n = n()
  ) %>%
  filter(n >= 10)  # Only bins with sufficient data

# Visualization with Smooth Curve
ggplot(win_rate_summary, aes(x = rating_diff_mid, y = win_percentage)) +
  geom_point(size = 3, color = "steelblue", alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "darkblue", linewidth = 1.2) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 100, y = 52, label = "50% = Even match", color = "red") +
  labs(
    title = "Empirical Win Rate vs Rating Difference",
    subtitle = "The S-shaped curve suggests a Logistic Regression approach",
    x = "Rating Difference (ELO Points)",
    y = "Win Percentage (%)"
  ) +
  scale_y_continuous(limits = c(40, 100), breaks = seq(40, 100, by = 10)) +
  theme_minimal()
```

### Descriptive Findings:

Baseline: When the rating difference is near 0, the win rate is approximately 50%, indicating a fair match.

Trend: The win rate increases steadily as the rating difference grows.

Linearity Check: Crucially, the relationship is not a straight line. It follows a sigmoidal (S-shaped) curve. This visual evidence confirms that a Logistic Regression model is the appropriate statistical tool.



### Logistic Regression Model Results

We now fit the Generalized Linear Model (GLM).

```{r}
# Fit Logistic Regression
logit_model <- glm(favorite_won ~ rating_diff, 
                   data = model_data, 
                   family = binomial(link = "logit"))

# Extract summary
tidy_results <- broom::tidy(logit_model)
kable(tidy_results, caption = "Logistic Regression Model Coefficients")
```

### Interpretation and Conclusion

1. Statistical Significance
The p-value for the rating_gap term is < 2e-16 (effectively zero).
Conclusion: We strongly reject the Null Hypothesis ($H_0$). There is overwhelmingly significant evidence that the rating difference predicts game outcomes.

2. Effect Size (Odds Ratio)
The coefficient estimate $\beta_1$ is approximately r round(coef(logit_model)[2], 4).
To interpret this, we calculate the Odds Ratio:
$$ \text{OR} = e^{\beta_1} \approx e^{0.0055} \approx 1.0055 $$

Practical Meaning: For every single ELO point difference, the odds of the favorite winning increase by 0.55%.
While small on a per-point basis, this compounds quickly:

A 100-point advantage increases the odds of winning by a factor of $\approx 1.73$.

A 200-point advantage increases the odds of winning by a factor of $\approx 3.00$.

3. Model Fit Visualization
Finally, we overlay the model's predicted probabilities curve against our observed data to check for fit.



```{r}
# Generate predictions for a smooth curve
grid_data <- data.frame(rating_diff = seq(0, 800, 1))

# Predict returns probabilities (0-1). We calculate Standard Error to show confidence.
preds <- predict(logit_model, newdata = grid_data, type = "link", se.fit = TRUE)

# Convert from Log-Odds back to Probability
grid_data$predicted_prob <- plogis(preds$fit) * 100
grid_data$lower_ci <- plogis(preds$fit - 1.96 * preds$se.fit) * 100
grid_data$upper_ci <- plogis(preds$fit + 1.96 * preds$se.fit) * 100

ggplot() +
  # Actual data points
  geom_point(data = win_rate_summary, aes(x = rating_diff_mid, y = win_percentage), 
             color = "steelblue", size = 3, alpha = 0.7) +
  # Confidence Interval Ribbon (New Addition)
  geom_ribbon(data = grid_data, aes(x = rating_diff, ymin = lower_ci, ymax = upper_ci), 
              fill = "red", alpha = 0.2) +
  # Model prediction curve
  geom_line(data = grid_data, aes(x = rating_diff, y = predicted_prob), 
            color = "darkred", linewidth = 1.2) +
  labs(
    title = "Model Fit: Predicted vs Observed Win Rates",
    subtitle = "Red Line = Logistic Model | Shaded Area = 95% Confidence Interval",
    x = "Rating Difference",
    y = "Win Percentage (%)"
  ) +
  theme_minimal()
```


The red prediction line tracks the observed data points extremely closely. The narrow confidence interval (shaded red area) indicates that our model is very precise due to the large sample size, confirming that rating difference is a highly reliable predictor of chess outcomes.





